---
title: "Week 2: Data Cleaning & Exploration"
output:
  word_document: default
  html_document: default
date: "2022-08-01"
---

# Step 0: Clean/inspect data


```{r}

Sea_urchins <- read.csv("Desktop/Design & Analyses/Datasets/Sea_urchins.csv")

summary(Sea_urchins)

# Check Datatypes 

result="hide"
fig.show ="hide"
```


```{r}
# Converting into factors 
Sea_urchins$Site <-factor(Sea_urchins$Site)
Sea_urchins$Area <-factor(Sea_urchins$Area)
Sea_urchins$Month <-factor(Sea_urchins$Month)
Sea_urchins$TD <- factor(Sea_urchins$TD)

result="hide"
fig.show ="hide"
```

```{r}
# Collapsing

library(tidyverse)
Sea_urchins$Area <- fct_collapse(Sea_urchins$Area, High = c("A" , "B"), Low = c("C","D"))
summary(Sea_urchins$Area)

# Areas with high and low fishing pressure

result="hide"
fig.show ="hide"
```

Step 0 Conclusion: summarized the data, converted to factors, collapsed the data


# Step 1: Finding the Outliers

```{r}
# Finding the outliers for GSI boxplot and dotchart

x <-Sea_urchins$GSI
boxplot(x)
dotchart(x)

result="hide"
fig.show ="hide"
```

```{r}
# Sensitivity Analysis

library(tidyverse)
Q3 <- quantile(Sea_urchins$GSI, probs = .75, na.rm = TRUE)
IQR <- IQR(Sea_urchins$GSI, na.rm = TRUE)
seaurchins1 <- subset(Sea_urchins$GSI, Sea_urchins$GSI < (Q3 + 1.5*IQR))
  
result="hide"
fig.show ="hide"
```

```{r}
boxplot(seaurchins1)

# Outliers have been removed

result="hide"
fig.show ="hide"
```

Step 1 Conclusion: made a boxplot and dotchart, found outliers, and removed outliers



# Step 2: Homogeneity of variance

```{r}
library(tidyverse)
Sea_urchins$Area <- fct_collapse(Sea_urchins$Area, High = c("A" , "B"), Low = c("C","D"))
summary(Sea_urchins$Area)

library(lattice)
bwplot(Sea_urchins$GSI ~ Sea_urchins$Area | Sea_urchins$Site)

result="hide"
fig.show ="hide"
```


```{r}
# Creating a residual plot vs fitted plot for TD
model <- lm(Sea_urchins$GSI ~ Sea_urchins$TD, data = Sea_urchins)

plot(model)

result="hide"
fig.show ="hide"
```

Step 2 Conclusion: we have equal variances within each group



# Step 3: Checking for Normality

```{r}
# Checking for normality 
x <-Sea_urchins$GSI

{hist(x, freq=FALSE) 
  curve(dnorm(x,mean=mean(x), sd=sd(x)), 
        add=TRUE, type = "l")}

# QQ Normality Plot
qqnorm(x)
qqline(x)

result="hide"
fig.show ="hide"
```

Step 3 Conclusion: data is not normally distributed




# Step 4: Are there lots of zeros in the data?

Does not apply because we are not using count/discrete data, we're using continuous data




# Step 5: Is there colinearity among the covariates?


```{r}

library(car)
model <- lm(Sea_urchins$GSI ~ Sea_urchins$GW + Sea_urchins$TD + Sea_urchins$TW, data = Sea_urchins)
# vif(model) --> returned an error "aliased coefficients in the model" so had to remove TW

# VIF of TW is highest (14.2) so removing that variable

model <- lm(Sea_urchins$GSI ~ Sea_urchins$GW + Sea_urchins$TD, data = Sea_urchins)
vif(model)

# VIF of GW and TD are now both less than 3 so this step is complete

result="hide"
fig.show ="hide"
```


Step 5 Conclusion: had to drop TW to reduce collinearity among the covariates




# Step 6: What are the relationships between x and y variables?

```{r}
# Plot response vs. each covariate

plot(Sea_urchins$GSI ~ Sea_urchins$GW)
plot(Sea_urchins$GSI ~ Sea_urchins$TD)

result="hide"
fig.show ="hide"
```

Step 6 Conclusions: GW and GSI show a positive non-linear relationship, as do TD and GSI



# Step 7: Should we consider interactions?

```{r}
# Coplot

coplot(GSI ~ TD | Month, data = Sea_urchins)

result="hide"
fig.show ="hide"
```

Step 7 Conclusion: interactions to be discussed in week 4


# Step 8: Are observations of the response variable independent?

```{r}
plot(Sea_urchins$GSI ~ Sea_urchins$Month)

boxplot(Sea_urchins$GSI ~ Sea_urchins$Month)

result="hide"
fig.show ="hide"
```

Step 8 Conclusions: measures do not appear to be repeated, observations appear independent 



# Would an ANOVA be an appropriate test to examine the research question?

- Unsure because we have a large sample size, but the data isn't normally distributed




