---
title: "Week 3 Assignment"
output: html_document
date: "2022-08-08"
---



```{r echo=T, results='hide', fig.show='hide'}
SU <- read_csv("Desktop/Design & Analyses/Datasets/Sea_urchins_cleaned.csv")

hist(SU$GSI) # Checking normality of GSI
bwplot(SU$GSI ~ SU$Size) # Checking homogeneity of variances of GSI by size class

```



# Log Transformation of GSI

```{r echo=T, results='hide', fig.show='hide'}
SU$GSI_log <- log(SU$GSI) # Transforming the GSI data with a log transformation

hist(SU$GSI_log) # Histogram to check for normality of our log-transformed data

qqnorm(SU$GSI_log) # QQ Plot to check for normality of our log-transformed data
qqline(SU$GSI_log)

shapiro.test(SU$GSI_log) # Shaprio-Wilk test to check for normality of our log-transformed data
```

```{r echo=T, results='hide', fig.show='hide'}
# Checking for homogeneity of variances of the log-transformed GSI of different size classes
library(lattice)
bwplot(SU$GSI_log ~ SU$Size)

# Creating a residual plot vs fitted plot
log.model <- lm(SU$GSI_log ~ SU$Size)

plot(log.model)

```


Is the assumption of normality still violated?
- Our log-transformed data still violates the assumption of normality 

Is the assumption of homogeneity of variances still violated?
- No, our variances are now homogeneous after log-transforming our data




# Square Root Transformation of GSI

```{r echo=T, results='hide', fig.show='hide'}
SU$GSI_sqrt <- sqrt(SU$GSI) # Transforming the GSI data with a square root transformation

hist(SU$GSI_sqrt) # Histogram to check for normality of our sqrt-transformed data

qqnorm(SU$GSI_sqrt) #  QQ Plot to check for normality of our sqrt-transformed data
qqline(SU$GSI_sqrt)

shapiro.test(SU$GSI_sqrt) # Shaprio-Wilk test to check for normality of our sqrt-transformed data

```

```{r echo=T, results='hide', fig.show='hide'}
# Checking for homogeneity of variances of the sqrt-transformed GSI of different size classes
library(lattice)
bwplot(SU$GSI_sqrt ~ SU$Size)

# Creating a residual plot vs fitted plot
sqrt.model <- lm(SU$GSI_sqrt ~ SU$Size)

plot(sqrt.model)
```


Is the assumption of normality still violated?
- GSI looks fairly normally distributed after being transformed but the Shapiro-Wilk test shows that the data is still not normally distributed

Is the assumption of homogeneity of variances still violated?
- No, our variances are now homogeneous after square-root transforming our data





# Subsetting data by size class IF WE NEED IT LATER
CS_log <- subset(SU, Size == "CS")
US_log <- subset(SU, Size == "US")
Small_US_log <- subset(SU, Size == "Small-US")




# Permutation test to test whether GSI differs among the size classes

```{r echo=T, results='hide', fig.show='hide'}
# Step 1: calculate test statistic on original data set

anova <- aov(SU$GSI ~ SU$Size)
summary(anova)
```


```{r echo=T, results='hide', fig.show='hide'}
# Step 2: run permutation

output <- rep(NA, 10000) # Empty vector (NA) to store t-values with length "replicates" as set above

for(i in 1:10000){ # Create a for loop
        SU$GSI_rando <- sample(SU$GSI, replace = FALSE) #randomize variable
       a <- aov(GSI_rando ~ Size, data = SU) # run ANOVA and store output in "a"
       output[i] <- a$statistic # extract F-value from "a" and store in output 
       }

dist <- hist(as.numeric(output)) #create a distribution of new F-values (i.e. our new null hypothesis)

```

I think you need to include a line of code that actually asks for the summary (e.g. summary(aov) 0 before you run the output[i] line




```{r echo=T, results='hide', fig.show='hide'}
# Step 3: calculate p-value of permutation test


```


Note: ANOVA is always a one-tailed test
P = 1-percentile

Conclusion: 





