---
title: "MB5003 Intro to R"
output: html_document
date: "2023-03-30"
---

```{r}
library(ggplot2)
library(tidyverse)
```


# Challenge No. 1

```{r}
(-2 + sqrt(10^2 - 4*2*3))/(2*log(9)*exp(2))
```


Import data

```{r}
landings_data <- read.csv("~/Desktop/MB5003 Fisheries Science/landings_data_clean.csv")

table(landings_data$Year)
table(landings_data$Gear)
table(landings_data$Species)
```

Data Transformation

```{r}
subset_gillnet <- landings_data %>% 
  filter(Gear == "gillnet", Year == 2005) %>% 
  mutate(Weight_kg = Weight_g/1000)

subset_gillnet %>% 
  summarise(total_weight = sum(Weight_kg))
```


# Challenge No. 2

Create a table that shows annual survivorship (S) and correlated mortality rate (Z) (Chapter 5 of Russ 2013)

```{r}
challenge_2 <- data.frame(survivorship_S = rev(seq(0, 1, length.out = 21)))

challenge_2
```

Use formula: Z = -ln(S)

```{r}
challenge_2 <- mutate(challenge_2, mortality_rate_Z = -log(survivorship_S))

challenge_2
```

Plotting

```{r}
# Plot year and length

ggplot(landings_data) +
  geom_point(aes(x = Year, y = Length_cm, color = Gear, shape = Gear))

ggplot(landings_data) +
  geom_boxplot(aes(x = Year, y = Length_cm, group = Year))
```

That graph sucks. Synthesize the data first, then plot...

```{r}
annual_landings <- landings_data %>% 
  mutate(Weight_kg = Weight_g/1000) %>% 
  group_by(Year, Gear) %>% 
  summarise(Annual_Landings_kg = sum(Weight_kg))
```

```{r}
ggplot(annual_landings, aes(x = Year, y = Annual_Landings_kg, color = Gear, shape = Gear)) +
  geom_point() +
  geom_line()
```


# Challenge No. 3

Use ggplot to plot the total landings or total catch (in kg) vs. year

```{r}
annual_landings %>% 
  group_by(Year) %>% 
  summarise(Total_catch_kg = sum(Annual_Landings_kg)) %>% 
  ggplot(aes(x = Year, y = Total_catch_kg)) +
  geom_point() +
  geom_line()
```

Calculate CPUE

```{r}
cpue_data <- landings_data %>% 
  mutate(Weight_kg = Weight_g/1000) %>% 
  group_by(Year, Trip_ID) %>% 
  summarise(Trip_CPUE = sum(Weight_kg)/mean(Effort_Hours)) %>% 
  group_by(Year) %>% 
  summarise(Median_CPUE_kg_hour = median(Trip_CPUE))
```

Plot CPUE

```{r}
ggplot(cpue_data, aes(x = Year, y = Median_CPUE_kg_hour)) +
  geom_point() +
  geom_line()
```

Calculate percent mature

```{r}
#Define m95, the length at which 95% of fish are mature
m95 = 15.9 

percent_mature <- landings_data %>% 
  mutate(Mature = Length_cm > m95) %>% 
  group_by(Year) %>% 
  summarise(Percent_Mature = (sum(Mature)/n())*100)

percent_mature
```

```{r}
ggplot(percent_mature, aes(x = Year, y = Percent_Mature)) +
  geom_bar(stat = "identity")
```









