---
title: "Debris Stats Tests"
output: html_document
date: "2023-01-10"
---

```{r}
library(tidyverse)
library(readxl)
library(plotrix)
macroplastics <- read_excel("~/Desktop/MB5001 Macroplastics Counts.xlsx")
plastics_mass <- read_excel("~/Desktop/MB5001 Macroplastics Mass.xlsx")
```



# ABUNDANCE

## Difference in abundance between sites: chi-square goodness of fit test

```{r}
chisq.test(mean_for_sites$mean, p = c(1/3, 1/3, 1/3))
```
Conclusion: there is a significant difference between the observed frequency of debris across the sites and the expected frequency (expected = there will be an equal abundance of debris across all sites)


## Difference in abundance between sampling heights: chi-square goodness of fit test

```{r}
chisq.test(mean_for_heights$mean, p = c(1/5, 1/5, 1/5, 1/5, 1/5))
```
Conclusion: there is a significant difference between the observed frequency of debris across the sampling heights and the expected frequency (expected = there will be an equal abundance of debris across all sampling heights)



# MASS

## Difference in mass of debris between sampling heights: kruskal-wallis test

```{r}
kruskal.test(Total_weight_kg ~ Tidal_line, data = plastics_mass)
```
Conclusion: there is not a significant difference in mass of debris between sampling heights



# MATERIAL



# CATEGORY/SOURCE















