---
title: "MB5370 Module 1: Introduction to Programming"
author: Katya Jagolta
Affiliation: James Cook University
output: html_document
date: "2023-02-22"
---

Workspace: 
~/MB5370 Techniques in Mar Sci 1/Module_01.R

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyr)
```


# Workshop 1

### 2.3.6 Getting started

Simple math:
```{r}
2+1
```

Returns a sequence of numbers from 1 to 30:
```{r}
1:30 
```


### 2.3.7 Variables, objects, and assignment

Assigning my age, my name, and then some simple math with age:
```{r}
age <- 26
first_name <- 'Katya' 
age + 1 
age + age
```

Errors:
Names can't start with a number and can't include special symbols or spaces!
(Made the following code comments in order to knit)

```{r}
# Errors
# 01_age <- 26
# !_age <- 26
# age katya <- 26
```

Uppercase vs. lowercase makes a difference:

```{r}
Age <- 45
age
Age
```

Exercise: sum numbers.

1. Add up numbers and assign to a name.
2. Name another number.
3. Add both together.

```{r}
apple <- 15+25.1+20.25
banana <- 17 
apple + banana 
```


### 2.3.9 Built-in functions.

1. Assign the value 26.7 to years_old.
2. Round up.
3. Round down.

```{r}
years_old <- 26.7 
round(years_old) 
floor(years_old) 
```

Exercise: use the paste function to make a statement of my age (rounded to 2 decimal places) using the objects I've made.

```{r}
name <- "Katya"
years_old <- 26.765
round(years_old, 2)

paste(name, "is", floor(years_old), "years old.")
```


### 2.3.10 Debugging

Misconceptions: Value of total is a spreadsheet would be 100, but in programming the variable maintains the value it was assigned (65) until reassigned.

```{r}
grade <- 55
total <- grade + 10
print (total) 
grade <- 90
print (total)
total <- grade + 10
print (total)
```

What will this code do?
The printed result should equal 10. 
```{r}
p <- 2 
z <- 5 
out <- p*z 
print (out) 
```

Error reports: 

```{r}
# out <- p * a
```
Error: object 'a' not found (because a is an undefined variable). The following should work because both p and z have already been defined:


```{r}
out <- p * z
out
```

Testing: the following will tell you whether x is a character and then whether it's numeric.
```{r}
x <- 1
is.character(x)
is.numeric(x)
```

Exercise: identify problems in this code...

1. Missing a comma.
2. Misspelled "my_quiz"
3. Function should be length()
```{r}
# my_quiz <- c("uno",
#             "dos",
#             "tres",
#             "quatro" 
#             "cinco")
# print (my_quis)
# str(my_quiz)
# len(my_quiz)
```

Fixed the code and testing it: creates and prints my_quiz, then tells you the data type and the length/number of elements.
```{r}
my_quiz <- c("uno",
             "dos",
             "tres",
             "quatro",
             "cinco")
print (my_quiz)
str(my_quiz)
length(my_quiz)
```





### 2.3.11 Data types

Exercise: try to produce the 4 data types (character, numeric, integer, logical)

```{r}
# Character:
char <- "testing"
class(char)

# Numeric:
num <- 7
class(num)

# Integer:
int <- as.integer(num) # Change num (7) from numeric to an integer
class(int)

# Logical:
log <- FALSE
class(log)
```


### 2.3.12 Data structures

Creating vectors of numeric elements, character elements, and a list. The final piece of code returns the 3rd element in the vector z.
```{r}
y <- c(1, 2, 3) 
z <- c("Sarah", "Tracy", "Jon") 
zz <- c(1, "Sarah", 3, "Jon")
yy <- list(1, "a", TRUE) 

class(y) 
class(z) 
class(zz) 
class(yy) 

z[3] 
```


### 2.3.13 Packages and libraries

Already have ggplot2, tidyr, and dplyr installed. 

```{r}
# install.packages("earthones")
# Unsuccessful! Package ‘earthones’ is not available for this version of R

# install.packages("redlistr")
# Success

# install.packages("terra")
# Success
```

Create a plot: Make data for the x-axis and y-axis, then plot y by x.

```{r}
x <- c(-1, -0.8, -0.6, -0.4, -0.2, 0, 0.2, 0.4, 0.6, 0.8, 1)
x
y <- x^3 
y
qqplot(x,y) 
```


### 2.3.14 Coding best practices

Headers are at top of script along with calling packages ggplot2 and tidyr. 



# Workshop 2

### 3.1.1 Learning about for-loops

1. Returns "one run" 4 times, one for each word
2. Returns each of the 4 words
3. Uses "value" instead of "word"; still returns each of the 4 words
4. sys.sleep() suspends execution of R expressions for a specified time interval, in this case 1 second to print each word - helpful if you want to watch & check what your code is doing as it runs. 

```{r}
for (value in c("My", "first", "for", "loop")) { print("one run")
} 

for (value in c("My", "second", "for", "loop")) { print(value)
} 

for (word in c("My", "third", "for", "loop")) { print(word)
} 

for (word in c("My", "fourth", "for", "loop")) { print(word)
  Sys.sleep(1)
} 
```

Custom for loop: For i, meaning every single element in my_list, this for loop will print "I am printing loop number i." and then will say "For loop complete" when it is complete!

```{r}
my_list <- 1:50
my_list
str(my_list)

for (i in my_list) {
  result <- i
  out <- paste("I am printing loop number:", result, ".")
  print(out)
  Sys.sleep(0.6)
}
print("For loop complete")
```

Adding a constant outside the loop: For each i, this for loop will print "I am printing loop number (i+500)." and then will say "For loop complete" when it is complete!

```{r}
add_me <- 500

for (i in my_list) {
  result <- i + add_me
  out <- paste("I am printing loop number:", result, ".")
  print(out)
  Sys.sleep(0.2)
} 
print("For loop complete") 
```

Nested for loop: For each round of the base loop (i), it runs all of the inside loop (j), producing something kind of like an outline in this case (1 abc, 2 abc, 3 abc, etc.).

```{r}
for (i in c(1,2,3,4,5)) {
  print(paste("I am base loop number: ", i))
  Sys.sleep(0.5)
  for (j in c("a","b","c")) {
    print(paste("... I am inside loop ", j))
  }
} 
print("Nested loop complete :)")
```

Exercise: for loop to list my favorite animals.

```{r}
numbers <- c(1,2,3,4,5)
animals <- c("tiger shark", "lobster", "blue crab", "shrimp", "cuttlefish")

for (i in numbers) {
  print(paste("My number", i, "favorite animal is a", animals[i]))
}
print("Thanks for coming to my TedTalk!")
```


### 3.1.2 If and if-else statements

1. Assigns the value 1 to x.
2. Says if x equals 1, it now equals 2 (reassigns x).
3. Confirming that x now does not equal 1 but equals 2.
4. If x is 1, print "x is: 1", otherwise print "We're running the else statement" and then print "x is: 2". 
```{r}
x <- 1

if (x == 1) {
  x <- 2
} 

x == 1 
x == 2 

if (x == 1) {
  print(paste("x is:", x))
} else {
  print("We're running the else statement")
  print(paste("x is:", x))
}
```

Exercise: make my own if else statement.

1. Says if name is Katya, the last name is Jagolta.
2. Prints my full name, Katya Jagolta.

```{r}
name <- "Katya"

if (name == "Katya") {
  last_name <- "Jagolta"
}
print(paste(name, last_name))
```

Changing the name:

1. Says if the name is Katya, the last name is Jagolta; otherwise, the last name is Kenney.
2. Reassigned name to "Kaysha", so it prints Kaysha Kenney.

```{r}
name <- "Kaysha"

if (name == "Katya") {
  last_name <- "Jagolta"
} else {
  last_name <- "Kenney"
} 
print(paste(name, last_name))
```

Changing name back to Katya see what that same if/else statement does: returns the correct name, Katya Jagolta.

```{r}
name <- "Katya"

if (name == "Katya") {
  last_name <- "Jagolta"
} else {
  last_name <- "Kenney"
}
print(paste(name, last_name))
```

Adding a new first name without an associated last name:

1. Says if name is Katya, last name is Jagolta; if name is Kaysha, last name is Kenney; otherwise, the last name is unknown.
2. Prints "Kyle... last name is unknown".

```{r}
name <- "Kyle" 

if (name == "Katya") {
  last_name <- "Jagolta"
} else if (name == "Kaysha"){
  last_name <- "Kenney"
} else {
  last_name <- "... last name is unknown"
} 
print(paste(name, last_name)) 
```


### 3.1.3 Functions

1. Formula to convert F to C (but temp_F isn't defined yet).
2. Convert 100 F to C = 38 C.
3. Convert 200 F to C = 93 C.

```{r}
# temp_C <- (temp_F - 32) * 5/9 
(100-32) * 5/9 
(200-32) * 5/9 
```

Make a function to convert F to C so we don't have to type this equation every time, then check that it works correctly:

```{r}
F_to_C <- function(temp_F){
  temp_C <- (temp_F - 32) * 5/9
  return(temp_C)
}

F_to_C(100)
F_to_C(200)
```

Exercise: write a function to convert celcius to fahrenheit, then test it.

```{r}
# temp_F <- temp_C * 9/5 + 32

C_to_F <- function(temp_C){
  temp_F <- temp_C * 9/5 + 32
  return(temp_F)
}

C_to_F(0) 
C_to_F(20) 
C_to_F(40)
```



### 3.1.4 Data frames

1. Make a dataframe.
2. Check the structure of the dataframe.
3. Convert col1 from numeric to factor (a category rather than a real continuous number).

(Col1 supposed to be factors at the end but still numeric; worked before but not working now, not sure why).

```{r}
my_dataframe <- data.frame(number = c(1,2,3), 
                           name = c("Katya", "Kaysha", "Kyle"),
                           logical = c(TRUE, TRUE, FALSE)) 
my_dataframe
str(my_dataframe)

as.factor(my_dataframe$no)
str(my_dataframe) 
```








