---
title: "MB5370 Module 4: Data Science"
author: Katya Jagolta
affiliation: James Cook University
output: html_document
date: "2023-05-02"
---

## Learning outcomes:
1. Learning plotting in detail using ggplot2
2. Learning reproducible science with Git/Github
3. Tidy data principles using tidyr
4. Mapping in R


# Workshop 1: Dataviz in R

```{r}
library(tidyverse)
```

## 2.5 Look at the mpg data

```{r}
mpg
```

## 2.6-2.7 Create a plot with the mpg data

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = class)) +
  labs(x = "Engine size (L)", y = "Fuel efficiency (mpg)")
```
This graph has engine size on the x-axis, fuel efficiency on the y-axis, and point color correlates to the class/type of car.


What happens if you map an aesthetic to something other than a variable name?

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = displ < 5)) +
  labs(x = "Engine size (L)", y = "Fuel efficiency (mpg)")
```

This changes the color of the points based on whether the engine size is greater than or less than 5 L. 


## 2.8 Troubleshooting

```{r}
# ggplot(data = mpg) 
# + geom_point(mapping = aes(x = displ, y = hwy)) 
# the = should be on top line

# Fixed:
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))  +
  labs(x = "Engine size (L)", y = "Fuel efficiency (mpg)")
```

## 2.9 Facet and panel plots

Facet the plot using a single variable (car class/type): facet_wrap(). Nrow() dictates how many rows of graphs there are (2). Ncol(2) would return two columns of graphs rather than 2 rows.

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_wrap(~ class, nrow = 2) +
  labs(x = "Engine size (L)", y = "Fuel efficiency (mpg)")
```


Facet the plot using more than one variable (# of cylinders and drive train type): facet_grid()

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ cyl) +
  labs(x = "Engine size (L)", y = "Fuel efficiency (mpg)")
```


## 2.10 Fitting simple lines

```{r}
ggplot(data = mpg) +
#   geom_point(mapping = aes(x = displ, y = hwy))  # points are messy
  geom_smooth(mapping = aes(x = displ, y = hwy)) + # line is nice & smooth
  labs(x = "Engine size (L)", y = "Fuel efficiency (mpg)")
```


Set the group aesthetic to a categorical variable to draw multiple objects:
```{r}
 ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))
```

Change the color of each line based on the drv value:
```{r}
 ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, color = drv))
```

Separate cars into 3 line types and point colors based on drive type: this looks very messy and is probably not the best way to display the data, but shows that each line comes from different data points (i.e. different drive train types).
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv, color = drv)) +
  geom_point(mapping = aes(x = displ, y = hwy, color = drv))  +
  labs(x = "Engine size (L)", y = "Fuel efficiency (mpg)")
```

More efficent way to write that code: put the x, y, and color in the ggplot() line rather than in both geom_point() and geom_smooth()
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, linetype = drv, color = drv)) + 
  geom_smooth() +
  geom_point() +
  labs(x = "Engine size (L)", y = "Fuel efficiency (mpg)")
```

To specify aesthetics in one layer but not another: geom_smooth uses the information within ggplot(); geom_point() uses that information as well as the additional aesthetics of changing color based on class
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point(mapping = aes(color = class)) +
  geom_smooth() +
  labs(x = "Engine size (L)", y = "Fuel efficiency (mpg)")
```

Can also specify different data for each layer: here we use filter() and (class == "subcompact") to plot only a subset of the data. This one shows points for all classes, but the smooth line only represents data for subcompact vehicles. I probably wouldn't use this exact plot, but this might be helpful with other datasets. 
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point(mapping = aes(color = class)) +
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE) + # se = FALSE leaves out the standard error shading
  labs(x = "Engine size (L)", y = "Fuel efficiency (mpg)")
```

Exercise:
1. Geoms for line chart, boxplot, histogram, and an area chart.

Line chart: not particularly helpful for this data
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = class)) +
  geom_line() +
  labs(x = "Engine size (L)", y = "Fuel efficiency (mpg)")
```


Boxplot:
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = class)) +
  geom_boxplot() +
  labs(x = "Engine size (L)", y = "Fuel efficiency (mpg)")
```

Histogram:
```{r}
ggplot(data = mpg, mapping = aes(hwy)) +
  geom_histogram() +
  labs(x = "Fuel efficiency (mpg)", y = "Frequency")
```

Area chart: not very useful for this data
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_area() +
  labs(x = "Engine size (L)", y = "Fuel efficiency (mpg)")
```


2 & 3. Predict what the code below will look like and then check.
I believe these two will look the same.

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```
And they do!


## 2.11 Transformations and Stats

Easy transformations & data summaries using the dataset 'diamonds' which includes price, carat, color, clarity, & cut of each diamond

The following shows that there are many more diamonds of high quality than low quality. By not specifying a y-value, ggplot counted the number of diamonds of each cut and plotted that.
```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut))
```

Can generally use geoms and stats interchangeably: here, stat_count() instead of geom_bar() does the same thing
```{r}
ggplot(data = diamonds) +
  stat_count(mapping = aes(x = cut))
```

Can override the default stat: stat = "identity" tells ggplot to use the raw values of variables rather than making default stats (a count or summary)
```{r}
demo <- tribble(
  ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551
)
demo

ggplot(data = demo) +
  geom_bar(mapping = aes(x = cut, y = freq), stat = "identity")
```


Can also override a default mapping from transformed variables to aesthetics. E.g.: display a proportion of the total rather than a count
```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, y = stat(prop), group = 1)) +
  labs(x = "Cut", y = "Proportion of all diamonds")
```

Can also show more information about the transformation: this shows the median, maximum, and minimum depth for each cut.
```{r}
ggplot(data = diamonds) +
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.min = min,
    fun.max = max, 
    fun = median
  )
```

Need to set group = 1 in proportion charts, otherwise this will happen: the data are (by default) grouped by cut, so the proportion of Fair in Fair is 100%, the proportion of Good in Good is 100%, etc. rather than the proportion of each cut out of all diamonds of all cuts.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = after_stat(prop)))
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = color, y = after_stat(prop)))
```

Group = 1 overrides the default behavior which is to group by the x variable (cut), so now the proportions of each cut are relative to all diamonds of all cuts.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = after_stat(prop), group = 1))

ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = after_stat(prop), fill = color, group = 1))
```

## 2.12 Positional adjustments

Change outline or fill of bars based on a variable: in this case there are separate colors for each cut
```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, color = cut))

ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = cut))
```

Change color to a different variable: clarity
```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = clarity))
```


Position adjustments allow you to customize plots in 3 ways:
- identity (raw data)
- fill (changes heights)
- dodge (forces ggplot2 to not put things on top of each other)
- jitter (slightly moves points so you can see them all)


Position = "identity": allows you to place each object exactly where it falls in the context of the graph
- Vital for point charts like scatter plots
- Horrible for bar plots - shows too much information when bar plots usually summarize information

```{r}
#To alter transparency (alpha)
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) +
  geom_bar(alpha = 1/5, position = "identity")

#To color the bar outlines with no fill color
ggplot(data = diamonds, mapping = aes(x = cut, colour = clarity)) + 
  geom_bar(fill = NA, position = "identity")
```

Position = "fill": stacks, but makes each set of stacked bars the same height to show proportions of each clarity level
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")
```

Position = "dodge": places overlapping objects directly beside one another
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")
```


Position = "jitter": slightly moves points so you can see them all; adds a small amount of random noise to each point to avoid overplotting when points overlap 
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), position = "jitter")
```


## 2.13 The layered grammar of graphics

Template for ggplot2:

ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(
     mapping = aes(<MAPPINGS>),
     stat = <STAT>, 
     position = <POSITION>
  ) +
  <FACET_FUNCTION>


## 2.14 Assignment: Plot Deconstruction

Import data: 
(Note: rate of homelessness is x out of 10,000 people)

Initial graph: 
- Did not have units
- Did not read easily 
- Said it represents the 10 states with the highest rates of homelessness but really showed 12
- Showed that the average rate of homelessness in the US decreased, while the article claimed that the rate had increased (article title: "America’s homeless population rises for first time in years")
- Is no longer found in the original article - maybe they realized it does not support the argument that the article is making



```{r}
homelessness <- read_excel("~/Desktop/MB5370 Techniques 1/MB5370 Module 4 Homelessness.xlsx")
```

Group by state and find the difference in homelessness rate from 2015 to 2017

```{r}
homelessness_summary <- homelessness %>% 
  group_by(State) %>% 
  mutate(Difference = `2017 Homelessness Rate` - `2015 Homelessness Rate`) %>% 
  filter(Type == "state")
```


```{r}
ggplot(data = homelessness_summary) +
  geom_bar(mapping = aes(x = reorder(State, -Difference), y = Difference, fill = "F8766D"), stat = "identity") +
  labs(x = "State", y = "Change per 10,000 people", title = "Change in homelessness from 2015 to 2017", caption = "Includes the 12 U.S. states with the highest rates of homelessness.
Values represent an increase or decrease in the number of homeless people per 10,000.") +
  theme_light() +
  theme(axis.text.x = element_text(angle = -45, hjust = 0), plot.caption = element_text(hjust = 0)) +
  geom_hline(aes(yintercept = -0.3, linetype = "U.S. Average"), color = "blue") +
  scale_linetype_manual(name = " ", values = c(1)) +
  guides(fill = FALSE)
```
```{r}
ggsave("Change in Homelessness.pdf", width = 7, height = 5)
```



