---
title: "Module 4 Workshop 4"
author: Katya Jagolta
affiliation: James Cook University
output: html_document
date: "2023-05-09"
---

# Workshop 4: Data Wrangling

```{r}
library(tidyverse)
```

## 5.3 Tibbles!

```{r}
iris
str(iris)
# Change data frame to a tibble:
as_tibble(iris)
```

```{r}
tibble(x = 1:5, y = 1, z = x ^ 2 + y)
# Call new variables to produce new column values
```


```{r}
x <- c(1:5)
y <- c(1)
z <- c(x^2+y)
data.frame(x, y, z) # Takes a lot more steps to make a dataframe than a tibble
```

```{r}
tb <- tibble(
  `:)` = "smile", 
  ` ` = "space", 
  `2000` = "number"
)
tb # Use "escape characters" (the little backtick marks) to separate code from labels
```


```{r}
# tribble: a transposed tibble to help you do data entry directly into a script. Column headings are defined by formulas (start with ~), and each data entry is put in a column, separated by commas

tribble(
  ~x, ~y, ~z,
  "a", 2, 3.6,
  "b", 1, 8.5
)
# Lays out small amounts of data in easy to read form
```

```{r}
# Exploring tibbles more 

tibble(
  a = lubridate::now() + runif(1e3) * 86400,
  b = lubridate::today() + runif(1e3) * 30,
  c = 1:1e3,
  d = runif(1e3),
  e = sample(letters, 1e3, replace = TRUE)
)
```

```{r}
install.packages("nycflights13")
library(nycflights13)
nycflights13::flights %>% 
  print(n = 10, width = Inf)
```

```{r}
options(tibble.width = Inf)
```

```{r}
# Pull out a single variable: can use $ to extract a full column (by name), or double bracket to pull out an exact row of data (by row position), or an exact cell value using [[,]]

df <- tibble(
  x = runif(5),
  y = rnorm(5)
)

# Extract by name
df$x
df[["x"]]

# Extract by row position
df[[1]]

# Extract by exact position
df[[2,2]]
```

```{r}
# Using pipes - have to use a '.' as a placeholder when using pipes to use base R functions
df %>% .$x
df %>% .[["x"]]
```

```{r}
# Tibbles won't do partial matching; if it doesn't match exactly, tibbles will give a warning

df <- tibble(
    xxx = runif(5),
    y = rnorm(5)
)
df$xxx
```

```{r}
df <- data.frame(abc = 1, xyz = "a") # Creating a dataframe 
df # Viewing dataframe
df$x # Call by name
df[, "xyz"] # Call by exact position
```

## 5.4 Importing data

```{r}
read_csv("a,b,c
1,2,3
4,5,6")

read_csv("The first line of metadata
  The second line of metadata
  x,y,z
  1,2,3", skip = 2)

read_csv("# A comment I want to skip
  x,y,z
  1,2,3", comment = "#")
```


```{r}
read_csv("1,2,3\n4,5,6", col_names = FALSE)
# \n is a shortcut for adding a new line; a break

# Add column names:
read_csv("1,2,3\n4,5,6", col_names = c("x", "y", "z"))

# Specify missing values as NA:
read_csv("a,b,c\n1,2,.", na = ".")
```


## 5.5 Tidying data using Tidyr

```{r}
library(tidyverse)
```


### 5.5.1 Tidy data

```{r}
# Table 1 is tidy, the rest are not:
table1
table2
table3

# Spread across 2 tibbles:
table4a # Cases
table4b # Population
```

Rules of tidy data:
1. Each variable must have its own column.
2. Each observation must have its own row.
3. Each value must have its own cell.

Even simpler rules:
1. Put each dataset in a tibble.
2. Put each variable in a column.

```{r}
# Ways you can work with/wrangle tidy data:

# Compute rate per 10,000: adds a new column called rate
table1 %>% 
  mutate(rate = cases / population * 10000)

# Compute cases per year
table1 %>% 
  count(year, wt = cases)

# Visualise changes over time
library(ggplot2)
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))
```


### 5.5.2 Spreading and gathering data tables

```{r}
# Change data in wide format to long format:
table4a <- table4a %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "cases")

table4b <- table4b %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "population")
```

```{r}
# Combine the 2 tables:
left_join(table4a, table4b)
```

```{r}
table2
# Need to pivot table2 wider since case #s and population #s are in the same column
table2 %>%
    pivot_wider(names_from = type, values_from = count)
```

### 5.5.3 Separating and uniting data tables

```{r}
table3
# Rate needs to be split into 2 columns; convenient that there's a break (/) between the two rates
table3 %>% 
  separate(rate, into = c("cases", "population"))
# By default, separate() will split values wherever there's a non-alphanumeric character

# To specify which character to separate by:
table3 %>% 
  separate(rate, into = c("cases", "population"), sep = "/")

# Both rates are listed as character types, but should be integers since they're actually numbers; ask separate() to convert them to better types:
table3 %>% 
  separate(rate, into = c("cases", "population"), convert = TRUE)

# Can separate/split a vector of integers:
table3 %>% 
  separate(year, into = c("century", "year"), sep = 2) # sep = 2 means splitting after the 2nd integer

# Opposite of separate(): unite() to combine multiple columns into a single column 
table5 %>% 
  unite(new, century, year, sep = "") # Add sep = "" so that we don't have the default underscore between the 2 sets of integers
```

### 5.5.4 Handling missing values

```{r}
# The return for the 4th quarter of 2015 is missing as NA:

stocks <- tibble(
  year   = c(2015, 2015, 2015, 2015, 2016, 2016, 2016),
  qtr    = c(   1,    2,    3,    4,    2,    3,    4),
  return = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)

stocks %>% 
  pivot_wider(names_from = year, values_from = return)
```

```{r}
# Set values_drop_na = TRUE in pivot_longer() to turn explicit missing values implicit

stocks %>% 
  pivot_wider(names_from = year, values_from = return) %>% 
  pivot_longer(
    cols = c(`2015`, `2016`), 
    names_to = "year", 
    values_to = "return", 
    values_drop_na = TRUE
  )
```

```{r}
# Making missing values explicit - very clear to you that they represent actual missing data values -  using complete(). This function takes a set of columns and finds all the unique combinations and then ensures the original dataset contains all of those values, including filling in explicit NA where needed

stocks
stocks %>% 
  complete(year, qtr)
```


```{r}
# The fill() function can be used to fill in missing values that were meant to be carried forward in the data entry process. It can take columns with missing values and carry the last observation forward (replace them with the most recent non-missing value)

treatment <- tribble(
  ~ person,           ~ treatment, ~response,
  "Derrick Whitmore", 1,           7,
  NA,                 2,           10,
  NA,                 3,           9,
  "Katherine Burke",  1,           4
)
treatment

treatment %>% 
  fill(person)
```

## 5.6 Learning relational data






