---
title: "Module 4 Workshop 5"
output: html_document
date: "2023-05-10"
---

# Workshop 5: Spatial data in R

```{r}
library(tidyverse)
library(sf) # simple features
library (terra) # for raster
library(tmap) # Thematic maps are geographical maps in which spatial data distributions are visualized
library(readr)
library(ggplot2)
```


## 6.6 Loading the spatial dataset

```{r}
dat <- read_csv("data/copepods_raw.csv")
dat
```


## 6.7 Data exploration

### 6.7.1 Check coordinates

```{r}
ggplot(dat) +
  aes(x = longitude, y = latitude, color = richness_raw) +
  geom_point()
# NOT a map since it doesn't have a projection! Just a scatterplot, but a nice quick way to look at spatial data
```

```{r}
# Plot richness by latitude 
ggplot(dat, aes(x = latitude, y = richness_raw)) + 
  stat_smooth() + 
  geom_point()
```


## 6.8 Getting going with maps

```{r}
# Use sf package (simple features) to turn point data into a spatially referenced dataframe 

sdat <- st_as_sf(dat, coords = c("longitude", "latitude"),
                 crs = 4326)
# st_as_sf: converts different data types to simple features
# dat: our original data 
# coords: gives the names of the columns that relate to the spatial coordinates (in order of X coord followed by Y coord)
# crs: coordinate reference system
```

Simple feature: like a shapefile in that it holds a lot of data in columns and rows, but it's spatially aware - extra columns regarding each rows position (in coordinates) and metadata about the coordinate reference system, the type of geometry (point) and so on.

## 6.9 Coordinate reference systems

- A datum = the reference point
- An ellipsoid = the lumpy spherical earth model
- Geographic coordinate reference system (GCS): combines the datum & ellipsoid to tell you coordinates of a location
- EPSG code: a short code for a specific coordinate reference system
- WKT (Well-known text): like the EPSG, a WKT can also be used to define a CRS but with a lot more detailed information

- WGS84: a common "one size fits all" GCS with EPSG = 4326 (input in the code above as crs = 4326)


```{r}
crs4326 <- st_crs(4326)
crs4326 # Look at the whole CRS
crs4326$Name # Pull out just the name of the crs
```

```{r}
crs4326$wkt # Crs in well-known text format
```


## 6.10 Feature collection (points)

```{r}
sdat
# Looks much like dat, but now has a geometry column too that stores coordinates (one point for each data row) 
```

## 6.11 Cartography

```{r}
plot(sdat["richness_raw"]) # Plots the richness column
plot(sdat) # Creates a panel for every variable in the dataframe
```


## 6.12 Thematic maps for communication

```{r}
# Using tmap:

tm1 <- tm_shape(sdat) +
  tm_dots(col = "richness_raw") # Plots dots of the coordinates with richness as the color scale

# Other plotting options: tm_polygons, tm_symbols, etc.
```

```{r}
tmap_save(tm1, filename = "Richness-map.png", 
          width = 600, height = 600)
```

The data above was a csv file with coordinate columns; below, the polygons in the copepod data are stored as shapefiles


## 6.13 Mapping spatial polygons as layers

### 6.13.1 Loading shapefiles

```{r}
aus <- st_read("data/spatial-data/Aussie/Aussie.shp")
shelf <- st_read("data/spatial-data/aus_shelf/aus_shelf.shp")
regions <- st_read("data/spatial-data/regions/regions.shp")
SST <- rast("data/spatial-data/MeanAVHRRSST/MeanAVHRRSST.grd")
```


```{r}
# Check out the data:
aus
shelf
regions
SST
```


```{r}
plot(SST)
```


### 6.13.2 Mapping your polygons

```{r}
# tmap makes it very straightforward to make a map of polygons:
tm_shape(shelf) +
  tm_polygons()
```

```{r}
tm_shape(shelf, bbox = sdat) +
  tm_polygons() +
  tm_shape(aus) + 
  tm_polygons() + 
  tm_shape(sdat) + 
  tm_dots()
```


```{r}
plot(SST)
```


## 6.14 Exploring t_map

```{r}
tmap_mode("plot") # Can toggle between plot (still map) and view (interactive map)

tm_shape(SST) +
  tm_raster(title = "SST", palette = "-RdBu") +
  
  tm_shape(regions) +
  tm_polygons(col = "region", palette = "Set2") +
  
  tm_shape(shelf, bbox = sdat) +
  tm_polygons(col = "turquoise") +
  
  tm_shape(aus) + 
  tm_polygons(col = "lightgreen") + 
  
  tm_shape(sdat) + 
  tm_dots(col = "richness_raw", palette = "-viridis")
```

```{r}
tmap_mode("plot") # Can toggle between plot (still map) and view (interactive map)

tm2 <- tm_shape(SST) +
  tm_raster(title = "SST", palette = "viridis", style = "cont") +
  
  tm_shape(aus) + 
  tm_polygons(col = "lightgreen") + 
  
  tm_shape(sdat) + 
  tm_dots(col = "richness_raw", palette = "YlOrRd", title = "Species richness")

tm2
```

# Cut out blue edges!!!!

```{r}
tmap_save(tm2, filename = "Richness-map2.png", 
          width = 2000, height = 2000)
```



